The effect of linking assumptions and number of response options on inferred scalar implicature rate

### 2018-04-30 writeup.tex
completed results section

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}

@@

The study was administered online through Amazon Mechanical Turk.
Participants were introduced to a set of cards with pictures of one or
two animals (Figure \ref{fig:stimuli}). They were told that a
blindfolded fictional character called Bob is going to guess what
animals are on the card. On each trial, participants saw a card as well
as a sentence representing Bob's guess. For example, they saw a card
with a cat on it and read the sentence \enquote{There is a cat on the
card.} The study ended after 24 trials. At the end participants
optionally provided demographic information. You can access and view
\href{**XXX\%20masoud\%20insert\%20link**}{the online study here}.

@@

The design had two main manipulaitons: the type of card and the type of
guess. There were two types of cards. Cards with only one animal on them
and cards with two animals. Animals were chosen from the following set:
cat, dog, and elephant There were three types of guesses: simple (e.g.
\emph{There is a cat}), conjunctive (e.g. \emph{There is a cat and a
dog}), and disjunctive (e.g. \emph{There is a cat or a dog}).

In each trial, the animal labels used in the guess and the animal images
on the card may have no overlap (e.g.~Image: dog, Guess: \emph{There is
a cat or an elephant}), a partial overlap (e.g.~Image: Cat, Guess:
\emph{There is a cat or an elephant}), or a total overlap (e.g.~Image:
cat and elephant, Guess: \emph{There is a cat or an elephant}). Crossing
the number of animals on the card, the type of guess, and the overlap
between the guess and the card results in 12 different possible trial
types. We chose 8 trial types (Figure \ref{fig:trials}), balancing the
number of one-animal vs.~two-animal cards, simple vs.~connective
guesses, and expected true vs.~false trials.

The study used five different types of measurements. 1. two-options
(true vs.~false) 2. two-options (wrong vs.~right) 3. three-options
(wrong, neither, right) 4. four-options (wrong, kinda wrong, kinda
right, right) 5. five-options (wrong, kinda wrong, neither, kinda right,
right).

@@

We are primarily concerned with the \enquote{rate of implicatures} in an
experimental study. Two trial types are predicted to include pragmatic
implicatures. First, trials where there are two animals on the card but
the fictional character guesses using the connective \emph{or}; for
example \enquote{cat or dog} when the card has both a cat and a dog on
it. We call such trials \enquote{scalar} trials. Second, trials where
there are two animals on the card but the character guesses only one;
for example \enquote{cat} when the card has a cat and a dog on it. We
call such trials \enquote{exhaustive}. In our assessment of implicature
rate we focus on these two types of trials.

We define \enquote{implicature rate} in two ways:

@@

\caption{Adults' two-alternative forced choice judgments in the connective guessing game.}\label{fig:binaryPlot}
\end{figure}

@@

\caption{Adults' three-alternative forced choice judgments in the connective guessing game.}\label{fig:ternaryPlot}
\end{figure}

@@

\caption{Adults' three-alternative forced choice judgments in the connective guessing game.}\label{fig:quaternaryPlot}
\end{figure}

@@

\caption{Adults' three-alternative forced choice judgments in the connective guessing game.}\label{fig:quinaryPlot}
\end{figure}

\begin{figure}
\centering
\includegraphics{writeup_files/figure-latex/unnamed-chunk-2-1.pdf}
\caption{}
\end{figure}

\begin{itemize}
\tightlist
\item
  make sure to break down based on whether participants had logical
  training or not.
\end{itemize}

@@

\begin{figure}
\centering
\includegraphics{writeup_files/figure-latex/unnamed-chunk-3-1.pdf}
\caption{}
\end{figure}

\begin{verbatim}
## Warning in (function (fn, par, lower = rep.int(-Inf, n), upper =
## rep.int(Inf, : failure to converge in 10000 evaluations
\end{verbatim}

\begin{verbatim}
## Warning in checkConv(attr(opt, "derivs"), opt$par, ctrl = control
## $checkConv, : Model failed to converge with max|grad| = 0.524298 (tol =
## 0.001, component 1)
\end{verbatim}

\begin{verbatim}
## Generalized linear mixed model fit by maximum likelihood (Laplace
##   Approximation) [glmerMod]
##  Family: binomial  ( logit )
## Formula: implicature ~ definition * response_type + trial_type + (1 +  
##     response_type | card) + (1 | participant)
##    Data: implicature_rate
## 
##      AIC      BIC   logLik deviance df.resid 
##   1783.4   1899.0   -871.7   1743.4     2380 
## 
## Scaled residuals: 
##     Min      1Q  Median      3Q     Max 
## -7.8815 -0.2261 -0.1198  0.2334 10.0887 
## 
## Random effects:
##  Groups      Name                    Variance Std.Dev. Corr             
##  participant (Intercept)             5.224316 2.28568                   
##  card        (Intercept)             0.008402 0.09166                   
##              response_typequaternary 0.084138 0.29007  -1.00            
##              response_typequinary    0.003720 0.06099  -0.79  0.81      
##              response_typeternary    0.044946 0.21201   0.90 -0.89 -0.67
## Number of obs: 2400, groups:  participant, 200; card, 3
## 
## Fixed effects:
##                                       Estimate Std. Error z value Pr(>|z|)
## (Intercept)                           -2.64555    0.43138  -6.133 8.63e-10
## definitionlow                         -0.02508    0.24943  -0.101    0.920
## response_typequaternary                3.47868    0.61328   5.672 1.41e-08
## response_typequinary                   3.44163    0.55426   6.209 5.32e-10
## response_typeternary                   0.29732    0.56967   0.522    0.602
## trial_typescalar                       0.85657    0.13861   6.180 6.41e-10
## definitionlow:response_typequaternary -6.08294    0.61009  -9.970  < 2e-16
## definitionlow:response_typequinary    -5.71913    0.50693 -11.282  < 2e-16
## definitionlow:response_typeternary    -1.21490    0.36931  -3.290    0.001
##                                          
## (Intercept)                           ***
## definitionlow                            
## response_typequaternary               ***
## response_typequinary                  ***
## response_typeternary                     
## trial_typescalar                      ***
## definitionlow:response_typequaternary ***
## definitionlow:response_typequinary    ***
## definitionlow:response_typeternary    ** 
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Correlation of Fixed Effects:
##                     (Intr) dfntnl rspns_typqt rspns_typqn rspns_typt
## definitinlw         -0.287                                          
## rspns_typqt         -0.724  0.202                                   
## rspns_typqn         -0.760  0.224  0.554                            
## rspns_typtr         -0.643  0.218  0.418       0.510                
## trl_typsclr         -0.218 -0.001  0.060       0.065       0.007    
## dfntnlw:rspns_typqt  0.214 -0.408 -0.330      -0.167      -0.101    
## dfntnlw:rspns_typqn  0.217 -0.492 -0.156      -0.309      -0.116    
## dfntnlw:rspns_typt   0.220 -0.675 -0.155      -0.170      -0.280    
##                     trl_ty dfntnlw:rspns_typqt dfntnlw:rspns_typqn
## definitinlw                                                       
## rspns_typqt                                                       
## rspns_typqn                                                       
## rspns_typtr                                                       
## trl_typsclr                                                       
## dfntnlw:rspns_typqt -0.098                                        
## dfntnlw:rspns_typqn -0.103  0.266                                 
## dfntnlw:rspns_typt  -0.036  0.298               0.349             
## convergence code: 0
## Model failed to converge with max|grad| = 0.524298 (tol = 0.001, component 1)
## failure to converge in 10000 evaluations
\end{verbatim}

@@

>>>>

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\theoremstyle{definition}
\newtheorem{example}{Example}
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}

@@

The study was administered online through Amazon Mechanical Turk.
Participants were introduced to a set of cards with pictures of one or
two animals (Figure \ref{fig:stimuli}). They were told that a
blindfolded fictional character called Bob is going to guess what
animals are on the card. On each trial, participants saw a card as well
as a sentence representing Bob's guess. For example, they saw a card
with a cat on it and read the sentence \enquote{There is a cat on the
card.} The study ended after 24 trials. At the end participants
optionally provided demographic information. We also asked participants
if they had any prior training in logic. You can access and view on the
study's \href{https://github.com/thegricean/si-paradigms/}{online
repository}.

@@

The study had two main manipulaitons within participants: the type of
card and the type of guess. There were two types of cards. Cards with
only one animal on them and cards with two animals. Animals were chosen
from the following set: cat, dog, and elephant There were three types of
guesses: simple (e.g. \emph{There is a cat}), conjunctive (e.g.
\emph{There is a cat and a dog}), and disjunctive (e.g. \emph{There is a
cat or a dog}). In each trial, the animal labels used in the guess and
the animal images on the card may have no overlap (e.g.~Image: cat,
Guess: \emph{There is an elephant}), a partial overlap (e.g.~Image: cat,
Guess: \emph{There is a cat or a dog}), or a total overlap (e.g.~Image:
cat and dog, Guess: \emph{There is a cat or a dog}). Crossing the number
of animals on the card, the type of guess, and the overlap between the
guess and the card results in 12 different possible trial types. We
chose 8 trial types (Figure \ref{fig:trials}), balancing the number of
one-animal vs.~two-animal cards, simple vs.~connective guesses, and
expected true vs.~false trials. Three trials were randomly selected from
each of the 8 trial-types, for a total of 24 trials. The order of these
24 trials was randomized as well.

Participants could derive implicatures in two trial types. First, the
trial type in which two animals were present on the card (e.g.~cat and
dog) but Bob guessed only one of them (e.g. \enquote{there is a cat}).
Such trials can have a literal interpretation (cat and possibly more) or
an exhaustive interpretation (only cat). We refer to them as
\enquote{exhaustive}. The second trial type with implicatures was the
one in which two animals were on the card (e.g.~cat and dog) and Bob
used a disjunciton (e.g.~cat or dog). These trials can have a literal
(inclusive) interpretation (e.g.~cat or dog or both), or an exclusive
interpretation (e.g.~cat or dog, not both). We refer to these trials as
\enquote{scalar}. The following four trial types were used as
experimental control: two trial types where there was no overlap between
the guess (e.g.~elephant) and the animal(s) on the card (e.g.~cat, cat
and dog); and two trial types where the animal(s) on the card were
correctly guessed. For example, if there was only a cat on the card, Bob
said \enquote{there is a cat} and if there was a cat and a dog, Bob said
\enquote{there is a cat and a dog}.

The study also had a between participant manipulation of the number of
response options in the forced choice task. Participants were randomly
assigned to one of four different conditions. The conditions differed
with respect to the number of response options: binary (wrong
vs.~right), ternary (wrong, neither, right), quaternary (wrong, kinda
wrong, kinda right, right), and quinary (wrong, kinda wrong, neither,
kinda right, right). We wanted to see if the number of response options
in the forced choice task would affect our estimate of the task's
\enquote{implicature rate}.

@@

@@

\caption{Proportion responses for the two-alternative (binary) forced choice judgments in the guessing game.}\label{fig:binaryPlot}
\end{figure}

Here we present the proportion that participants chose the response
options in each of the 8 trial types of the binary, ternary, quaternary,
and quinary tasks. Figure \ref{fig:binaryPlot} shows the proportion of
\enquote{right} and \enquote{wrong} responses in the binary task.
Starting from the leftmost column, participants considered a guess
\enquote{wrong} if the guessed animal was not on the card. Moving to the
second column, participants considered a guess \enquote{right} if the
animal on the card was mentioned. However, if only one of the two
animals on the card was mentioned (exhaustive trials), 14\% of the times
participants considered the guess \enquote{wrong}. Moving to the third
column, if a conjunction of animals was guessed while only one animal
was on the card, participants considered the guess to be
\enquote{wrong}. If a conjunction of animals was guessed and both
animals were present on the card, all participants considered the guess
to be \enquote{right} as expected. Moving to the forth column, if a
disjunction of animals was guessed and only one of the animals was on
the card, participants considered the guess to be \enquote{right} almost
all the time. However, if both animals were present (scalar trials),
38\% of the times participants considered the guess to be
\enquote{wrong}.

@@

\caption{Proportion responses for the three-alternative (ternary) forced choice judgments in the guessing game.}\label{fig:ternaryPlot}
\end{figure}

Figure \ref{fig:ternaryPlot} shows the proportion of \enquote{right},
\enquote{neither}, and \enquote{wrong} responses in the ternary task.
Similar to the binary task, participants considered a guess wrong when
the mentioned animal was not on the card. They considered the guess
\enquote{right} when the mentioned animal was on the card. However, in
exhaustive trials when the fictional character only guessed one of the
two animals on the card, participants considered the guess
\enquote{wrong} 8\% of the time and neither wrong nor right 12\% of the
time. If a conjunction of animals was guessed and only one animal was
present on the card, participants considered the guess \enquote{wrong}.
As expected, when a conjunction was used and both animals were present,
participants considered the guess \enquote{right}. Similarly,
participants considerd the guess \enquote{right} when a disjunction was
used and only one of the animals was on the card. However, in scalar
trials that both animals were on the card and a disjunction was
gueassed, participants judged the guess \enquote{wrong} 23\% of the time
and \enquote{neither} 11\% of the time.

@@

\caption{Proportion responses for the four-alternative (quatenary) forced choice judgments in the guessing game.}\label{fig:quaternaryPlot}
\end{figure}

Figure \ref{fig:quaternaryPlot} shows the proportion of \enquote{right},
\enquote{kinda right}, \enquote{kinda wrong}, and \enquote{wrong}
responses in the quaternary task. Similar to the results seen
previously, the control trials turned outas expected. Participants
considered a guess \enquote{wrong} if the animal guessed was not on the
card and \enquote{right} if it was the only animal on the card. If a
conjunctin of animals was guessed and both animals were on the card the
guess was \enquote{right}. However, when only one of the animals on the
card was guessed (exhaustive trials), participants judged the guess
\enquote{wrong} 2\% of the time, \enquote{kinda wrong} 5\% of the time,
and \enquote{kinda right} 66\% of the times. Perhaps surprisingly, when
a conjunction was used and only one of the animals was on the card,
participants considered the guess \enquote{wrong} most of the time, but
they often considered it \enquote{kinda wrong} or even \enquote{kinda
right}. This suggests that perhaps participants considered a notion of
partially true or correct statement in our experimental setting.
Disjunctive guesses with one or two animals on the card showed similar
response patterns with participants choosing the \enquote{kinda right}
and \enquote{right} options most of the time. When both animals were on
the card with a disjunctive guess (scalar trials), participants judged
the guess \enquote{wrong} 6\% of the time, \enquote{kinda wrong} 12\% of
the time, and \enquote{kinda right} 43\% of the times.

@@

\caption{Proportion responses for the five-alternative (quinary) forced choice judgments in the guessing game.}\label{fig:quinaryPlot}
\end{figure}

Finally, Figure \ref{fig:quinaryPlot} shows the proportion of
\enquote{right}, \enquote{kinda right}, \enquote{neither},
\enquote{kinda wrong}, and \enquote{wrong} responses in the quinary
task. Since the results for the control trials were identical to
previous tasks, we do not repeat them here. In exhaustive trials where
two animals were on the card and only one of them was guessed,
participants chose \enquote{kinda right} the majority of times (72\%).
Again perhaps surprisingly, when only one animal was on the card and the
guess was a conjunction, responses were equally split among
\enquote{wrong}, \enquote{kinda wrong}, and \enquote{kinda right}
responses. With disjunctive guesses, partitipants were more likely to
choose \enquote{right} and \enquote{kinda right} options. When only one
animal was on the card, participants considered the disjunctive guess as
\enquote{right} more often. When both animals were on the card (scalar
trials), participants judged the disjunctive guess as \enquote{kinda
right} 52\% of the time.

Comparing the response patterns in binary to quinary tasks (Figures
\ref{fig:binaryPlot} to \ref{fig:quinaryPlot}), we can observe that in
implicature trials, participants are less likely to choose the endpoints
of the scale (i.e. \enquote{wrong} and \enquote{right}) as they are
given more intermediate options. This raises the possibilty that the
strict \enquote{wrong/right} responses in the tasks with fewer options
(binary and ternary) were not accurately reflecting participant
judgments of implicature trials and pushed their judgments to the
extreme ends. More generally, the data suggest that participant
judgments in contexts that give rise to implicatures fall within the
intermrediate level of the response scale and forced choice tasks with
few options (especially binary tasks) may risk depicting judgments more
extreme than they actually are.

@@

The goal of this study is to see if the estimated \enquote{implicature
rate} in the experimental task is affected by the linking assumptions
and the number of response options available in the task. Our analysis
in this section focuses on these three elements. Two trial types are
predicted to include pragmatic implicatures. First, trials where two
animals were on the card but the fictional character guessed with a
disjunction; for example \enquote{cat or dog} when the card has both a
cat and a dog on it. We called such trials \enquote{scalar}. Second,
trials where there were two animals on the card but the character
guessed only one; for example \enquote{cat} when the card had a cat and
a dog on it. We called such trials \enquote{exhaustive}. In our
assessment of implicature rate, we focus on these two types of trials.

\begin{figure}
\centering
\includegraphics{writeup_files/figure-latex/implicatureRatePlot-1.pdf}
\caption{}
\end{figure}

\begin{itemize}
\tightlist
\item
  Was logical training a significant predictor of \enquote{implicature
  rate}?
\end{itemize}

@@

A simple visual inspection of our results suggests that any variant of
the Traditional Linking Hypothesis will yield unsatisfying empirical
coverage with regards to participants' behavior in truth value judgment
tasks. Recall that according to this hypothesis, scalar implicature is
conceptualized as a binary, categorical affair: an implicature is either
\enquote{calculated,} or it isn't. This assumption has implications for
how we approach analysis of variation in behavior on a truth value
judgment task; for example, why did the majority of respondents in the
binary condition of our experiment answer \enquote{Right} to an
utterance of cat or dog when the card had a cat and a dog on it?

On the Traditional story, we are forced to say that a) not all
participants calculated the implicature; or that b) some participants
who calculated the implicature did not choose the anticipated response
(i.e. \enquote{Wrong}) due to some other cognitive reflex which
\enquote{overrode} the implicature; or some mixture of (b) and (c). We
might similarly posit that one or both of these factors underlie the
variation in the ternary, quatenary, and quinary conditions (e.g.~why
were participants roughly split between \enquote{Right} and
\enquote{Kind of right} when the utterance was cat or dog and the card
had a cat and a dog?). There is an established cottage industry of
pragmatics research which approaches the problem of behavioral variation
in exactly these ways, exploring the factors which \enquote{license}
implicature calculation or \enquote{suppress} its anticipated behavioral
expression (c.f. \_\_), but the best we can hope for on this approach is
an analysis which traces the general qualitative patterns in the data
(and we should stress that even in this respect, the industry has yet to
fully deliver the goods!).

The Alternative Linking Hypothesis contrasts with the Traditional
Linking Hypothesis in that the former is rooted in a quantitative
formalization of pragmatic competence which provides us a continuous
measure of pragmatic reasoning. {[}recap RSA model here? Maybe show how
it gets the right basic qualitative pattern, i.e. \(P_L\)(a and
b\textbar{} \enquote{a or b}) \(<\) \(P_L\)(a \textbar{} \enquote{or});
\(P_L\)(a and b\textbar{} \enquote{a or b}) \(<\) \(P_L\)(b \textbar{}
\enquote{or}) {]}

Following Degen \& Goodman (2014), we proceed on the assumption that
behavior on sentence verification tasks, such as truth value judgment
tasks, is best modeled as a function of an individual's mental
representation of a cooperative interlocutor (\(S_1\) in the language of
RSA). For a given utterance u and an intended communicated meaning w,
\(S_1\)(u \textbar{} w) outputs a conditional probability of u given w.
For example, in the binary condition of our experiment where a
participant evaluated cat or dog when there were both animals on the
card, the participant has access to the mental representation of \(S_1\)
and hence to the \(S_1\) conditional probability of hearing the
utterance cat or dog given a dog and cat card \(S_1\)(cat or dog
\textbar{} cat and dog).

One desideratum of a linking hypothesis given this view of pragmatic
competence is to transform this probability value into a categorical
output (e.g. \enquote{Right}/''Wrong'' in the case of the binary
condition). This can be achieved by positing that participants
\enquote{filter out} low-probability utterances given an intended
communicated meaning. We model a responder, R, who in the binary
condition responds \enquote{Right} to an utterance u in world w just in
case \(S_1\)(u \textbar{} w) exceeds some probability threshold
\(\theta\):

R(u, w, \(\theta\)) = \enquote{Right} iff \(S_1\)(u \textbar{} w) \(>\)
\(\theta\) = \enquote{Wrong} otherwise

In the experiment conditions where there are more than two choices, we
model the range of possible behavioral responses for R with the
introduction of intermediate probability thresholds. For example, in the
ternary condition, R(u, w, \(\theta_1\) , \(\theta_2\)) is
\enquote{Right} iff \(S_1\)(u \textbar{} w) \textgreater{} \(\theta_1\)
and \enquote{Neither} iff \(\theta_1\) \textgreater{} \(S_1\)(u
\textbar{} w) \textgreater{} \(\theta_2\). To fully generalize the model
to our five experimental conditions, we say that R takes as its input an
utterance u and a world state w and a number of threshold variables
dependent on a variable c, corresponding to the experimental condition
in which the participant finds herself (e.g.~the range of possible
responses available to R).

Given c = \enquote{ternary}

R(u, w, \(\theta_1\) , \(\theta_2\))\\
= \enquote{Right} iff \(S_1\)(u \textbar{} w) \(>\) \(\theta_1\) =
\enquote{Neither} iff \(\theta_1\) \(>\) \(S_1\)(u \textbar{} w) \(>\)
\(\theta_2\)\\
= \enquote{Wrong} otherwise

Given c = \enquote{quatenary}

R(u, w, \(\theta_1\) , \(\theta_2\), \(\theta_3\))\\
= \enquote{Right} iff \(S_1\)(u \textbar{} w) \(>\) \(\theta_1\) =
\enquote{Kinda Right} iff \(\theta_1\) \(>\) \(S_1\)(u \textbar{} w)
\(>\) \(\theta_2\)\\
= \enquote{Kinda Wrong} iff \(\theta_2\) \(>\) \(S_1\)(u \textbar{} w)
\(>\) \(\theta_3\) = \enquote{Wrong} otherwise

Given c = \enquote{quinary}

R(u, w, \(\theta_1\) , \(\theta_2\), \(\theta_3\). \(\theta_4\))\\
= \enquote{Right} iff \(S_1\)(u \textbar{} w) \(>\) \(\theta_1\)
=\enquote{Kinda Right} iff \(\theta_1\) \(>\) \(S_1\)(u \textbar{} w)
\(>\) \(\theta_2\) = \enquote{Neither} iff \(\theta_2\) \(>\) \(S_1\)(u
\textbar{} w) \(>\) \(\theta_3\)\\
= \enquote{Kinda Wrong} iff \(\theta_3\) \(>\) \(S_1\)(u \textbar{} w)
\(>\) \(\theta_4\) = \enquote{Wrong} otherwise
